---
title: "Using extract_transcripts in drawProteins"
author:
- name: "Dr Paul Brennan"
affiliation: 
- "Centre for Medical Education, School of Medicine, Cardiff University, 
    Cardiff, Wales, United Kingdom"
email: BrennanP@cardiff.ac.uk
package: drawProteins
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Using drawProteins}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r load_packages, eval = TRUE, echo=FALSE}
library(BiocStyle)
library(drawProteins)
library(ggplot2)
library(knitr)
opts_chunk$set(comment=NA,
                fig.align = "center",
                out.width = "100%",
                dpi = 100)
```

# Introducing extract_transcripts() in drawProteins
Many proteins are present as alternate transcripts where the same gene is 
produces alternative forms of the protein through differential mRNA splicing or
post-translational cleavage. 

These are detailed in UniProt. When they are extracted by the UniProt API, it
gives lists of alternative forms followed by lists of features. In order to 
plot each protein and the appropriate features, these need to be separated in 
our dataframe. This is done using the `extract_transcripts()` function. 

This Vignette shows how this works and gives an example. 

The workflow using extract_transcripts() is:

1. to provide one or more Uniprot IDs
2. get a list of features from the Uniprot API
3. run `extract_transcripts()` to generate a new dataframe
4. draw the chains and features as desired

Steps 1 and 2 are illustrated in drawProteins Vignette so only step3 and the 
visualisation of step 4 will be shown here. 


# Making a new dataframe with each transcript separated
The NFkappaB transcription factor family contains two proteins that are present
in two forms. The dataframe obtained from Uniprot is contained in the
drawProtein package as "five_rel_data" and can be loaded using the `data()` 
function. 

When loaded this has 320 obs of 9 variables and will plot five chains as
shown by checking the `max(five_rel_data$order)` function. 

To plot all the transcripts, a new dataframe is produced using the 
`extact_transcripts()` function. The new dataframe is called prot_data and 
has 430 obs of 9 variables and will plot seven chains as shown by checking
the `max(prot_data$order)` function.

```{r load_NFkappaB_data, fig.height=10, fig.wide = TRUE}
# load up data for five NF-kappaB proteins
data("five_rel_data")
max(five_rel_data$order)
# returns 5

# use extract_transcripts() to create a new data frame
prot_data <- extract_transcripts(five_rel_data)
max(prot_data$order)
# returns 7
```

Now, let's check out the chains for the two objects for comparison purposes.

```{r check_chains, fig.height=10, fig.wide = TRUE}
p1 <- draw_canvas(five_rel_data)
p1 <- draw_chains(p1, five_rel_data)
p1 <- p1 + ggtitle("Five chains plotted")

p2 <- draw_canvas(prot_data)
p2 <- draw_chains(p2, prot_data)
p2 <- p2 + ggtitle("Seven chains plotted")

p1
p2
```

The appropriate domains and phosphorylation sites can be drawn correctly. 

```{r draw_domains_and_phospho, fig.height=10, fig.wide = TRUE}
p2 <- draw_domains(p2, prot_data)
p2 <- draw_phospho(p2, prot_data, size =8) 
p2

```

Note that the names of the different transcripts are the same so it's wise to 
use the option customize the labels.

```{r draw_canvas_and_chains, fig.height=8, fig.wide = TRUE}
p2 <- draw_canvas(prot_data)
p2 <- draw_chains(p2, prot_data,
            fill = "lightsteelblue1", 
            outline = "grey",
            labels = c("p105",
                        "p105",
                        "p100", 
                        "p100",
                        "Rel B",
                        "c-Rel", 
                        "p65/Rel A",
                        "p50",
                        "p52"),
            label_size = 5)
p2 <- draw_phospho(p2, prot_data, size = 8, fill = "red")
p2 + theme_bw()
```

# Session info
Here is the output of `sessionInfo()` on the system on which this document was
compiled:
```{r session_Info, echo=FALSE}
sessionInfo()
```
